### 2018-01-17 submission/manuscript.tex
grammar corrections before a lab reading

\textbf{Results.} At 35 cycles there were significant differences
between HiFi DNA polymerase for fecal samples (P-value \textless{}
0.0001). These HiFi dependent differences in the number of OTUs could be
identified as early as 20 cycles in the mock communities (P-value =
0.002). Chimera prevalence varied by HiFi DNA polymerase and these
differences were still observed after chimera removal using VSEARCH.
Further, the chimera prevalence had a strong positive correlation with
the number of OTUs observed in a sample that was also not changed by
chimera removal with VSEARCH.\\
\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
number of OTUs, common diversity metrics that incorporate this value
could give artificially inflated numbers due to higher undetected
chimeras. So, when designing 16S rRNA gene survey studies it is
important to consider both the cycle number and the type of HiFi DNA
polymerase that will be used since it can increase or decrease the
number of OTUS that are observed.

@@

Over recent years there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, a lot of the attention of
this standardization push has been invested in investigating ways that
16S rRNA gene surveys can be made more replicatable (Lauber et al.,
2010; Salter et al., 2014; Song et al., 2016; Gohl et al., 2016).
Although 16S rRNA gene sequencing has been much maligned for introduced
bias, many of these same considerations also affect metagenomic
sequencing (Nayfach \& Pollard, 2016; Costea et al., 2017). Between the
two approaches similar bias considerations include, but are not limited
to, preservation media, storage conditions, DNA extraction kit, PCR, and
sequece library preparation. Thus what is identified as a source of bias
in 16S rRNA gene sequencing for these overlapping considerations will
also influence metagenomic sequencing results.

@@

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that differences, based on DNA polymerase, in the number of
OTUs generated were not easily removed using the authors chosen
bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that they may since
many of these HiFi DNA polymerase come from different families and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerase
introduce noticeable biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples
using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
polymerase. First, we tested if we could identify differences in the
number of OTUs between the differen HiFi DNA polymerase for both the
fecal and mock samples. Next, since there were differences based of HiFi
DNA polymerase, we examined if a community measure such as the
Bray-Curtis index would be affected. After assessing these community
measures we then examined if the different HiFi DNA polymerase had
different per base error rates as well as chimera prevalence. Finally,
since differences in chimera prevalence, based on HiFi DNA polymerase,
could not be completely removed using bioinformatic approaches we tested
whether the chimera prevalence correlated well with total number of
observed OTUs.

@@

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
If the HiFi Taq had a specific activation time that was different then 2
minutes that was used instead. For Kappa and Q5 the protocol previously
published by Gohl and colleagues was used (Gohl et al., 2016). For
Phusion the company defined conditions were used but the same extension
time as that used for Accuprime and Platinum was used.

@@

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009). The
protocol followed was similar to what has been previously publsihed
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use VSEARCH instead
of UCHIME for chimera detection and the use of the OptiClust algorithmn
instead of average neighbor for Operational Taxonomic Unit (OTU)
generation (Edgar et al., 2011; Rognes et al., 2016; Westcott \&
Schloss, 2017). Sequence error was determined using the seq.error
command on mock samples after chimera removal and classification to the
RDP to remove non-bacterial sequences (Schloss et al., 2009; Cole et
al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
graphing was completed using the tidyverse package (v 1.1.1) and colors
selected using the viridis package (v 0.4.0) (Garnier, 2017; Wickham,
2017). The total number of OTUs were analyzed using an ANOVA with a
tukey post-hoc test. For the fecal samples the data was normalized to
each individual by cycle number to account for the biological variation
between different people. Bray-Curtis matrices were generated using
mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The
distance matrix data was analyzed using PERMANOVA with the vegan package
(v 2.4.4) (Oksanen et al., 2017) and kruskal-wallis tests within R. For
both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curits indicies were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). From these observations we wanted to next analyze
potential reasons as to why some of these differences may have occured.
First, analysis of general sequence error rate, number of sequences with
an error, and base subsitution were assessed in the mock community for
each Taq. After assessing these errors, the total number of chimeras was
determined after sequence processing. The fecal samples were analyzed at
4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the
mock community samples were analysed at 3 levels, 1000, 5000, 10000.

@@

\textbf{\emph{The Number of OTUs is Dependent on HiFi Taq Used:}} After
normalization by individual, for each cycle number, we observed that for
fecal samples the number of OTUs identified was dependent upon the HiFi
DNA polymerase used and this difference increased as the depth of
sub-sampling increased {[}Figure 1{]}. Lower cycle numbers (15-20)
resulted in less differences between HiFI DNA polymerase while cycle
numbers of 25, 30, and 35 had larger clearer defined differences
{[}Figure 1{]}. Only 35 cycles had HiFi Taq differences that were
significantly different at all sub-sampling levels (P-value \textless{}
0.0001) {[}Table S1{]}. At sub-sampling depth of 5000 or higher 25 and
30 cycles had HiFi DNA polymerase differences (P-value \textless{} 0.05)
{[}Table S1{]}. Using a Tukey post-hoc test only at 35 cycles were
significant differences found to be mainly driven by Platinum being
different than all other Taq across sub-sampling depth (P-value
\textless{} 0.05) {[}Table S2{]}.

@@

\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
Number:}} Overall, there was very little difference between each
respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
samples and this was consistent across the different sub-samplings used
{[}Figure 3{]}. Two exceptions to the low differences between 5-cycle
increments can be found. The first that there seems to be large
differences for fecal samples between 20x vs.~25x that is robust against
sub-sampling depth {[}Figure 3A-B{]}. Second, for the mocks, where data
is available, there appears to be a similar large difference detected
between 20x and 25x {[}Figure 3D{]}. Regardless, by the time PCR cycles
reach 25x there does not seem to be large differences in the community
{[}Figure 3{]}.

@@

\textbf{\emph{Sequence Error is Dependent on both Taq and Cycle Number
Used:}} Differences by HiFi Taq in the median average per base error
varied without a clear pattern across sub-sampling depth {[}Table S5{]}.
Generally, the highest values were for the Kappa HiFI DNA polymerase
{[}Figure 4A{]}. Sub-sampling depth seems to have little effect on this
rate with both 5000 and 10000 sub-sampled sequences showing similar
results {[}Figure 4B-C{]}. There were small differences between the
various HiFi Taq at lower cycle number but larger differences at higher
cycle number with Platinum having large differences between all other
HiFi DNA polymerase {[}Figure 4B-C and Table S6{]}.

The total seqeunces with at least one error had multiple differences at
different cycle numbers and sub-sampling depth driven by large
differences in Accuprime\textsuperscript{TM} and Platinum versus the
other HiFi Taq tested {[}Figure S1 \& Table S7 \& S8{]}. Although
Accuprime\textsuperscript{TM} had the lowest per base error rate it had
either the largest or second largest number of sequences with at least
one error regardless of cycle number or sub-sampling depth {[}Figure
S1{]}. Investigation of whether there were HiFi DNA polymerase dependent
effects on base subsitution found that there was no clear bias and this
was indpendent of sub-sampling depth {[}Figure S2-S4{]}. Further, the
variation in subsitution error seems to reduce as the sub-sampling depth
increases {[}Figure S2-S4{]}.

\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi Taq
Dependent:}} After chimera removal using VSEARCH and removal of
sequences that did not classify as bacteria we assessed the percentage
of sequences that were still chimeric within our mock community. At all
levels of sub-sampling and cycle number there were significant
differences between the HiFi DNA polymerase used (P-value \textless{}
0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast majority of
these differences were driven by Platinum being different then all other
HiFi DNA polymerase across cycle number and sub-sampling depth {[}Table
S10{]}. Generally, across sub-sampling depth and cycle number
Accuprime\textsuperscript{TM} had the lowest chimera prevalence of all
the HiFi DNA Polymerase regardless of whether pre.cluster or VSEARCH had
been used {[}Figure 5{]}.

For all Taqs a positive correlation was observed between chimeric
sequences and number of OTUs, with this correlation being strongest for
Accuprime, Platinum and Phusion HiFi DNA Polymerase {[}Figure 6{]}. In
general, the R\textsuperscript{2} value between the number of OTUs and
chimeric sequences did not change from the use of pre.cluster and
VSEARCH {[}Figure 6{]}. Taken together this data suggests that a strong
correlation exists between the number of OTUs and the prevalence of
chimeric sequences. Kappa had the highest per base error rate and the
lowest correlations between the number of OTUs and chimeric sequences
across sub-sampling depth {[}Figure 4 \& 6{]}.

@@

Our observations build upon previous studies (Gohl et al., 2016) by
showing that even different HiFi DNA Polymerase have significant
differences in the number of OTUs and that the changes to total OTUs
correlate with chimeras not removed after sequence processing {[}Figure
1-2 \& 5{]}. This is important since many diversity metrics rely on the
total number of OTUs as part of their calculations and changes to the
total number of OTUs could drastically change the results as well as the
findings. Although the attention has mostly been on standardizing and
improving collection and extraction methods (Salter et al., 2014) our
observations show that independent of this consideration HiFi DNA
polymerase can have a noticeable affect on the OTUs generated that can
be found across sub-sampling depth and PCR cycle number {[}Figure
2-4{]}. These differences were observed in high biomass samples, where
biases introduced by such components like kit contamination have less of
an effect, suggesting that these differences may be exacerbated in low
biomass samples.

Although we did not observe strong differences, based on cycle number,
using the Bray-Curtis index the data suggests that there may be
differences between 15 cycles and higher cycle numbers, such as 30x,
that are commonly used. What is most interesting is that there was no
difference within individuals between corresponding 5 cycle increments
(e.g.~15 to 20, 20 to 25, etc.). Conversely, this may be due to low
power and on observation there does seem to be a trend that 20x and 25x
communities are very different {[}Figure 3{]}. This finding, in
conjunction with the PERMANOVA results, suggest that cycle number can
change bacterial community calculations but that these differences are
minimal once at least 25 cylces are reached. Increasing the sub-sampling
depth, for some DNA polymerase may reduce some of these observed
community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was very strongly correlated with the number of
OTUs and this value is relied upon heavily for many different downstream
community metric calculations. However, Bray-Curtis analysis with
PERMANOVA showed few differences based on DNA polymerase. Since it is
possible that many of the increased number of OTUs, generated as cycle
number increases, are not highly abundant that the Bray-Curtis index is
able to successfully downweight these respective OTUs (Minchin, 1987).
So, choice of downstream diversity metric could be an important
consideration to mitigate these observed changes due to high chimera
prevalence in HiFi DNA polymerase such as Platinum.

>>>>

\textbf{Results.} At 35 cycles there were significant differences
between HiFi DNA polymerase for fecal samples (P-value \textless{}
0.0001). These HiFi dependent differences in the number of OTUs could be
identified as early as 20 cycles in the mock communities (P-value =
0.002). Chimera prevalence varied by HiFi DNA polymerase and these
differences were still observed after chimera removal using VSEARCH.
Additionally, the chimera prevalence had a strong positive correlation
with the number of OTUs observed in a sample and was also not changed by
chimera removal with VSEARCH.

\textbf{Conclusions.} Due to the impact of HiFi DNA polymerase on the
number of OTUs, common diversity metrics that incorporate this value
could give artificially inflated numbers due to higher undetected
chimeras. So, when designing 16S rRNA gene survey studies it is
important to consider both the cycle number and the type of HiFi DNA
polymerase that will be used since it can increase or decrease the
number of OTUS that are observed.

@@

Over recent years there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, a lot of the attention of
this standardization push has been invested in investigating ways that
16S rRNA gene surveys can be made more replicatable (Lauber et al.,
2010; Salter et al., 2014; Song et al., 2016; Gohl et al., 2016).
Although 16S rRNA gene sequencing has been much maligned for introduced
bias, many of these same considerations also affect metagenomic
sequencing (Nayfach \& Pollard, 2016; Costea et al., 2017). Between the
two approaches similar bias considerations include, but are not limited
to, preservation media, storage conditions, DNA extraction kit, PCR, and
sequence library preparation. Thus what is identified as a source of
bias in 16S rRNA gene sequencing for these overlapping considerations
will also influence metagenomic sequencing results.

@@

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that differences, based on DNA polymerase, in the number of
OTUs generated were not easily removed using the authors chosen
bioinformatic pipeline (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families and may intrinsically have different error rates that cannot be
completely removed with modifications (Ishino \& Ishino, 2014). In this
study we critically examine if any of five different HiFi DNA polymerase
introduce noticeable biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

We amplified the 16S rRNA gene in both fecal and mock community samples
using either Accuprime, Kappa, Q5, Phusion, or Platinum HiFi DNA
polymerase. First, we tested if we could identify differences in the
number of OTUs between the differen HiFi DNA polymerase for both the
fecal and mock samples. Next, since there were differences based on HiFi
DNA polymerase, we examined if a community measure such as the
Bray-Curtis index would be affected. After assessing these community
measures we then examined if the different HiFi DNA polymerase had
different per base error rates as well as chimera prevalence. Finally,
since differences in chimera prevalence, based on HiFi DNA polymerase,
could not be completely removed using bioinformatic approaches we tested
whether the chimera prevalence correlated well with total number of
observed OTUs.

@@

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
If the HiFi Taq had a specific activation time that was different then 2
minutes, the specified activation was used instead. For Kappa and Q5 the
protocol previously published by Gohl and colleagues was used (Gohl et
al., 2016). For Phusion the company defined conditions were used but the
same extension time as that used for Accuprime and Platinum was used.

@@

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009). The
protocol followed was similar to what has been previously publsihed
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithmn instead of average neighbor for Operational Taxonomic Unit
(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
2016; Westcott \& Schloss, 2017). Sequence error was determined using
the seq.error command on mock samples after chimera removal and
classification to the RDP to remove non-bacterial sequences (Schloss et
al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number, to account for the
biological variation between different people. Bray-Curtis matrices were
generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
15000. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests
within R. For both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curits indicies were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons as to why some of these differences may have occured.
First, analysis of general sequence error rate, number of sequences with
an error, and base subsitution were assessed in the mock community for
each DNA polymerase. After assessing these errors, the total number of
chimeras was determined after sequence processing. For the community
based measures, the fecal samples were analyzed at 4 different
sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock
community samples were analysed at 3 levels, 1000, 5000, 10000.

@@

\textbf{\emph{The Number of OTUs is Dependent on HiFi DNA Polymerase
Used:}} After normalization by individual, for each cycle number, we
observed that for fecal samples the number of OTUs identified was
dependent upon the HiFi DNA polymerase used and this difference
increased as the depth of sub-sampling increased {[}Figure 1{]}. Lower
cycle numbers (15-20) resulted in less differences between HiFI DNA
polymerase while cycle numbers of 25, 30, and 35 had larger clearer
defined differences {[}Figure 1{]}. Only 35 cycles had HiFi DNA
polymerase differences that were significantly different at all
sub-sampling levels (P-value \textless{} 0.0001) {[}Table S1{]}. At
sub-sampling depth of 5000 or higher 25 and 30 cycles had HiFi DNA
polymerase differences (P-value \textless{} 0.05) {[}Table S1{]}. Using
a Tukey post-hoc test only at 35 cycles were significant differences
found to be mainly driven by Platinum being different than all other DNA
polymerases across sub-sampling depth (P-value \textless{} 0.05)
{[}Table S2{]}.

@@

\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
Number:}} Overall, there was very little difference between each
respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
samples and this was consistent across the different sub-samplings used
{[}Figure 3{]}. Two exceptions to the low differences between 5-cycle
increments can be found. The first, that there seems to be large
differences for fecal samples between 20x vs.~25x that is robust against
sub-sampling depth {[}Figure 3A-B{]}. Second, for the mocks, where data
is available, there appears to be a similar large difference detected
between 20x and 25x {[}Figure 3D{]}. Regardless, by the time PCR cycles
reach 25x there does not seem to be large differences in the community
between the same sample {[}Figure 3{]}.

@@

\textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
and Cycle Number Used:}} Differences by HiFi DNA polymerase in the
median average per base error varied without a clear pattern across
sub-sampling depth {[}Table S5{]}. Generally, the highest values were
for the Kappa HiFI DNA polymerase and the pre-cluster step and chimera
removal by VSEARCH had little effect on this rate {[}Figure 4{]}.
Generally, there were small differences between the various HiFi DNA
polymerase at lower cycle numbers and larger differences at higher cycle
number with Platinum having large differences between all other HiFi DNA
polymerase {[}Figure 4B-C and Table S6{]}.

The total seqeunces with at least one error had multiple differences at
different cycle numbers and was mostly alleviated by the use of the
pre.cluster step {[}Figure S1{]}. Major differences before this
pre.cluster step were driven by large differences in
Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
polymerase tested {[}Figure S1 \& Table S7 \& S8{]}. Although
Accuprime\textsuperscript{TM} had the lowest per base error rate it had
the largest number of sequences with at least one error, regardless of
cycle number or sub-sampling depth {[}Figure S1{]}. However, this
increased number of sequences with an error can be drastically lowered
with existing bioinformatic approaches {[}Figure S1{]}. Investigation of
whether there were HiFi DNA polymerase dependent effects on base
subsitution found that there were generally no biases in the types of
subsitution made {[}Figure S2{]}.

\textbf{\emph{Chimeric Sequences Correlate with OTUs and are HiFi DNA
Polymerase Dependent:}} After chimera removal using VSEARCH and removal
of sequences that did not classify as bacteria we assessed the
percentage of sequences that were still chimeric within our mock
community. At all levels of sub-sampling and cycle number there were
significant differences between the HiFi DNA polymerase used (P-value
\textless{} 0.05) {[}Table S9{]}. Using a Dunn's post-hoc test the vast
majority of these differences were driven by Platinum being different
then all other HiFi DNA polymerase across cycle number and sub-sampling
depth {[}Table S10{]}. Generally, across sub-sampling depth and cycle
number Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all the HiFi DNA Polymerase regardless of whether pre.cluster or
VSEARCH had been used {[}Figure 5{]}.

For all DNA polymerase, a positive correlation was observed between
chimeric sequences and number of OTUs, with this correlation being
strongest for Accuprime, Platinum and Phusion HiFi DNA Polymerase
{[}Figure 6{]}. In general, the R\textsuperscript{2} value between the
number of OTUs and chimeric sequences did not change from the use of
pre.cluster and VSEARCH {[}Figure 6{]}. Taken together this data
suggests that a strong correlation exists between the number of OTUs and
the prevalence of chimeric sequences. Additionally, Kappa had the
highest per base error rate and the lowest correlations between the
number of OTUs and chimeric sequences across sub-sampling depth
{[}Figure 4 \& 6{]}.

@@

Our observations build upon previous studies (Gohl et al., 2016) by
showing that even different HiFi DNA polymerase have significant
differences in the number of OTUs and that the changes to total OTUs
correlate with chimeras not removed after sequence processing {[}Figure
1-2 \& 5{]}. This is important since many diversity metrics rely on the
total number of OTUs as part of their calculations and changes to the
total number of OTUs could drastically change the results, as well as
the findings. Although the attention has mostly been focused on
standardizing and improving collection and extraction methods (Salter et
al., 2014) our observations show that independent of this consideration
HiFi DNA polymerase can have a noticeable affect on the OTUs generated
that can be found across sub-sampling depth and PCR cycle number
{[}Figure 2-4{]}. These differences were observed in high biomass
samples, where biases introduced by such components like kit
contamination may have less of an effect, suggesting that these
differences may be exacerbated in low biomass samples.

Although we did not observe strong differences, based on cycle number,
using the Bray-Curtis index the data suggests that there may be
differences between 15 cycles and higher cycle numbers, such as 30x,
that are commonly used. Additionally, there was no difference within
individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
to 25, etc.). Conversely, this may be due to low power and on
observation there does seem to be a trend that 20x and 25x communities
are very different from each other {[}Figure 3{]}. This finding, in
conjunction with the PERMANOVA results, suggest that cycle number can
change bacterial community calculations but that these differences are
minimal once 25 cylces are reached. Increasing the sub-sampling depth,
for some DNA polymerase, may reduce some of these observed community
differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
and this value is relied upon heavily for many different downstream
community metric calculations. However, Bray-Curtis analysis with
PERMANOVA showed few differences based on DNA polymerase. Since it is
possible that many of the increased number of OTUs, generated as cycle
number increases, are not highly abundant allowing the Bray-Curtis index
to be able to successfully downweight these respective OTUs (Minchin,
1987). So, choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in HiFi DNA polymerase.
